Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-03 Manuscript/manuscript.tex
Corrected section on binding affinites and oxygen moieties cations bind to.

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{Kucerka2011}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with the experimental values from \cite{ferreira13}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

\begin{table}
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    The changes of the head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  \todo{Add temperatures for POPC and DPPC experiments} \\
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[tb]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of POPC bilayer as a function of NaCl concentrations
    from simulations with different force fields together with experimental data for DPPC~\cite{akutsu81} and POPC~\cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
  \todo{Add temperatures to the experimental data.} \\
\end{figure}

@@

\subsection{Affinity and binding cites to POPC membrane}

\begin{table}
  \caption{The relative surface excess and the probabilities of contacts between bound calcium and different
    oxygen moieties in POPC from different simulation models. The CaCl$_2$ concentration of the ions in water
    is 350mM in the analyzed systems. \label{tab:binding}}
  \begin{tabular}{l|c c c}
    model                 & $\Gamma_{Ca}^w$ (nm$^{-2}$)   & $P^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $P^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.07 \pm 0.01 $            &  99\%   &    25\%    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $            &  -       &     -       \\
    Lipid14/Dang          &  $ 0.3 \pm 0.03  $           &  -        &    -       \\
  \end{tabular}
\end{table}

The binding affinities of Ca$^{2+}$ ions to POPC bilayer in different simulation 
models were quantified by calculating the relative surface excess from Eq.~\ref{surfexcess} 
using the bulk concentrations determined from the density profiles in Fig.~\ref{fig:cacl-dens}. 
The relative surface excess values, shown in Table~\ref{tab:binding},
were calculated from simulations with the same, $350\,\mathrm{mM}$, concentration of CaCl$_2$ with 
respect to water, i.e., the concentration before solvating the lipids in experiments.
As expected from the changes of the lipid headgroup order parameters in Fig. \ref{fig:delta_ordPar_CaCl}, 
the relative surface excess for the ECC-lipids model, $\Gamma_{\rm Ca}^w = 0.07 \pm 0.01 \rm{nm}^{-2}$,
is significantly smaller than for the Lipid14 model with \AA{}qvist ions, $\Gamma_{\rm Ca}^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_{\rm Ca}^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) 
calculated using density profiles in Fig.~\ref{fig:cacl-dens}
is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which predict a specific binding of sodium on 
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
%
%This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
%which show the density profile of ions and water along the membrane normal. 
%
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipids model.

\begin{figure}[tb]
  \centering
  \includegraphics[height=20.0cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    densities of~\ce{Cl^-} and \ce{Na^{+}} are divided by 2, and
    densities of phosphate group resp. water are divided by 5 resp. 200. 
    The molar concentration of ions in water is 350~mM in all systems
    presented here. 
    }
\end{figure}

The probabilities for a bound \ce{Ca^{2+}} ion to interact with
different oxygen moieties of POPC 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by counting the contacts within the distance of $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}. 
%
%SAMULI: I commented out the average number of contacts. All the essential information
%for this is in the probabilities.
%
%In contrast to sodium, the dominant contribution
%to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
%% Numbers from simulation with 346mM conc. CaCl2 (OPC3)
%The average number of contacts per lipid 
%between \ce{Ca^{2+}} at $290\,\mathrm{mM}$ concentration and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are considered 
%%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
%
% Numbers from simulation with 980/845mM conc. CaCl2 (OPC3)
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
The results in Table \ref{tab:binding} show that almost all (99\%) of the 
bound Ca$^{2+}$ ions are in contact with the phosphate oxygens,
with two thirds (67\%) are interacting with only phosphate oxygens
and the rest, one third (32\%), interacts also with the carbonyl oxygens. 
Thus, the interactions between
calcium ions with only carbonyl oxygens are very unlikely (1\%), but significant
amount of the ions interact with both, phosphate and carbonyl oxygens.
The individual lipids or acyl chains were not identified in the analysis,
thus the simultaneous ion contacts with both oxygen moieties may be inter or intra molecular
and occur with the carbonyls in the either of the acyl chains.
The most likely contacts between Ca$^{2+}$ ions and phosphate oxygens is visualized with
the probability isodensity contours in Fig.~\ref{fig:volmaps}.
While the higher concentrations of \ce{CaCl2} naturally increased the amount of contacts per lipid,
%for $850\,\mathrm{mM}$ concentration of \ce{CaCl2}
%the average number of contacts per lipid is increased to
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $, 
the distribution of contacts between phosphate and carbonyl oxygens was not affected.
%and density profiles of ions and water along the membrane normal in Fig.~\ref{fig:cacl-dens}.

\begin{figure}[tb]
  \centering
  \includegraphics[width=6.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Contours of probability isodensities of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-lipids simulation.
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}
  SAMULI: I am not sure if this would be needed anymore.}
\end{figure}

Even thought \ce{Na+} ions bind to POPC bilayer with a very low probability,
it was possible to analyze the distribution of contacts from simulations with a large
sodium concetration of~1\,M.
%the average number of contacts per lipid with any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%this number decreases to 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are counted 
%it is even lower, 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
The results show that \ce{Na+} ions bind only to the phosphate oxygens 
of POPC with the probablity of 55\% and only to the carbonyl oxygens with the
probability of 20\%, leaving the probability of 25\% for the
simultaneous binding to both.%, phosphate and carbonyl oxygens.  
%This is in line with the calculated negative value of 
%relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
%their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
%suggesting that the populations of bound sodium cations 
%naturally decrease with the decreasing density of water.

In conclusion, the results suggest that the calcium ions specifically bind to phosphate
oxygens, while occasionally interacting also with carbonyls. This is in good agreement
with the previous conclusions from several experimental and theoretical
studies~\cite{hauser76,hauser78,herbette84,cevc90,binder02}, but
suggests lower relative binding affinity to carbonyls than the previous MD simulation
studies~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Also the sodium ions bind most likely to the phosphate oxygens, while the probability
of binding purely to carbonyls is also significant, in contrast to calcium ions.
The physiological relevance of the sodium binding details is, however, uncertain
due to the very weak binding affinity.

>>>>

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{Kucerka2011}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with the experimental values from \cite{ferreira13}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of headgroup order parameters and P-N vector orientation as a function of
    cationic surfactant (dihexadecyldimethylammonium) in POPC bilayer from simulations
    and experiments \cite{scherer89}.
  }
\end{figure}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    The changes of the head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  \todo{Add temperatures for POPC and DPPC experiments} 
  %\todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    %the value of $g_3$ order parameter of DPPC was -0.214 
    %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    %at the C-2 segments of both fatty acyl chains.'' }
\end{figure}

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    The changes of the head group order parameters of POPC bilayer as a function of NaCl concentrations
    from simulations with different force fields together with experimental data for DPPC~\cite{akutsu81} and POPC~\cite{altenbach84}.
    Ion concentrations in x-axis are calculated as in Fig.~\ref{fig:delta_ordPar_CaCl}.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC1000mMNaClfiles}.
  }
  %\todo{Add also the POPC result to the figure. Ref.\cite{altenbach84} gives for POPC at 1M of NaCl $\Delta S^\alpha=(6.1-5.5)*0.00784=0.004704$.} \\
  \todo{Add temperatures to the experimental data.} 
\end{figure}

@@

\subsection{Structure and affinity of cation binding to POPC membrane}

\begin{table}[tb!]
  \caption{Relative surface excess of calcium with respect to water, 
  and the ratios of the amount of calcium cations 
  bound to phosphate or carbonyl moieties 
  to the amount of such cations bound to both moieties in a POPC bilayer 
  from various simulation models. 
  The CaCl$_2$ concentration of the ions in water before adsorption to the phospholipid bilayer
  is 350mM in the analyzed systems. \label{tab:binding}}
  \todo{Add columns with concentrations of calcium -- bulk/molar fraction.}
  \todo{Changing symbol $P$ to $R$ or $r$ might reflect better that it is rather a ratio, and not that much a probability.}
  \begin{tabular}{l|c c c}
    model                 & $\Gamma_{Ca}^w$ (nm$^{-2}$)   & $P^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $P^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.06 \pm 0.01 $            &  99\%   &    25\%    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $            &  -       &     -       \\
    Lipid14/Dang          &  $ 0.3 \pm 0.03 $            &  -        &    -       \\
    Lipid14/ECC-ions      &  $ \simeq 0.4   $            &  -        &    -       \\
  \end{tabular}
\end{table}

Binding affinities of Ca$^{2+}$ ions to a POPC bilayer in different simulation 
models were quantified by calculating the relative surface excess of calcium to water molecules, $\Gamma_{\rm Ca}^w$, from Eq.~\ref{surfexcess}.
%using the bulk concentrations determined from the density profiles in Fig.~\ref{fig:cacl-dens}. 
The values of $\Gamma_{\rm Ca}^w$, shown in Table~\ref{tab:binding},
were calculated from simulations with the same $350\,\mathrm{mM}$ concentration of CaCl$_2$ with 
respect to water, i.e., the concentration before solvating the lipids in experiments.
As expected from the changes of the lipid headgroup order parameters in Fig. \ref{fig:delta_ordPar_CaCl}, 
the relative surface excess for the ECC-lipids model, $\Gamma_{\rm Ca}^w = (0.06 \pm 0.01) \rm{nm}^{-2}$,
is significantly smaller than for the Lipid14 model with \AA{}qvist ions, $\Gamma_{\rm Ca}^w = (0.13 \pm 0.01) \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_{\rm Ca}^w = (0.30 \pm 0.03) \rm{nm}^{-2}$,
or with ECC-ions,  $\Gamma_{\rm Ca}^w \simeq 0.4 \rm{nm}^{-2}$,
where the large uncertainity arises from a too small simulation box (Fig.~\ref{fig:cacl-dens}).
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) 
%calculated using density profiles in Fig.~\ref{fig:cacl-dens}
is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = (-0.11 \pm 0.01) \rm{nm}^{-2}$ 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to the most of the available lipid force fields,
which predict a specific binding of sodium to
PC lipid bilayers \cite{catte16}.
%\todo{JOE: Add references to NaCl simulation papers that show peaks.
%Add a simple note that there is NBFIX trick in Charmm36 (?).} 
%SAMULI: I think that it is enough to cite NMRlipids II here, which shows all the results
%
%This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
%which show the density profile of ions and water along the membrane normal. 
%
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipids model.

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    densities of~\ce{Cl^-} and \ce{Na^{+}} are divided by 2, and
    densities of phosphate group resp. water are divided by 5 resp. 200. 
    The molar concentration of ions in water is 350~mM in all systems
    presented here. 
    }
\end{figure}

The ratio of the amount of calcium cations 
bound to phosphate or carbonyl moieties 
to the amount of such cations bound to both moieties in a POPC bilayer 
from various simulation models
%The probabilities for a bound \ce{Ca^{2+}} ion to interact with
%different oxygen moieties of POPC 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
were analyzed by counting the contacts within the distance of $0.3 \, \mathrm{nm}$, 
as done previously in Ref.~\citenum{javanainen17}. 
%
%SAMULI: I commented out the average number of contacts. All the essential information
%for this is in the probabilities.
%
%In contrast to sodium, the dominant contribution
%to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
%% Numbers from simulation with 346mM conc. CaCl2 (OPC3)
%The average number of contacts per lipid 
%between \ce{Ca^{2+}} at $290\,\mathrm{mM}$ concentration and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are considered 
%%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
%
% Numbers from simulation with 980/845mM conc. CaCl2 (OPC3)
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
The results in Table \ref{tab:binding} show that almost all (99\%) of the 
bound Ca$^{2+}$ ions are in contact with the phosphate oxygens,
with two thirds (67\%) interacting with only phosphate oxygens
and the rest, one third (32\%), interacting also with the carbonyl oxygens. 
Thus, the interactions between
calcium ions with only carbonyl oxygens are very unlikely (1\%), however, a significant
amount of the ions interacts with both, phosphate and carbonyl, oxygens moieties.
The individual lipids or acyl chains were not distinguished in the analysis,
hence, a simultaneous contacts of ions with both oxygen moieties may be inter or intra molecular,
and may occur with carbonyls in either of the acyl chains.
The most likely interactions between Ca$^{2+}$ ions and phosphate oxygens are visualized with
the probability density isocontours in Fig.~\ref{fig:volmaps}.
While the higher concentrations of \ce{CaCl2} naturally increased the amount of contacts per lipid,
%for $850\,\mathrm{mM}$ concentration of \ce{CaCl2}
%the average number of contacts per lipid is increased to
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $, 
the distribution of contacts between phosphate and carbonyl oxygens was not affected.
%and density profiles of ions and water along the membrane normal in Fig.~\ref{fig:cacl-dens}.

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Isocontours of probability density of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-lipids simulation.
      The probability density was evaluated around a single lipid, 
      after its strucutral alignment using only phosphate group. 
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}
  SAMULI: I am not sure if this would be needed anymore.}
\end{figure}

Even though \ce{Na+} ions do not specifically bind to a POPC bilayer,
they still interact with its oxygen moieties. 
%the average number of contacts per lipid with any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
%whereas if only phosphate oxygen atoms are considered 
%this number decreases to 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
%and if only carbonyl oxygen atoms are counted 
%it is even lower, 
%$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
The results  from simulation at $1\,$M \ce{NaCl} concentration show 
that 55\% of \ce{Na+} ions interact with only phosphate oxygens of POPC 
and 20\% with only carbonyl oxygens,
and the rest, 25\%, is interacting with both.
%, phosphate and carbonyl oxygens.  
%This is in line with the calculated negative value of 
%relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
%their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
%suggesting that the populations of bound sodium cations 
%naturally decrease with the decreasing density of water.

In conclusion, the results suggest that calcium ions specifically bind to phosphate
oxygens occasionally interacting also with carbonyls. This is in good agreement
with the previous conclusions from several experimental and theoretical
studies~\cite{hauser76,hauser78,herbette84,cevc90,binder02}, but
suggests lower relative binding affinity to carbonyls than the previous MD simulation
studies~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Sodium ions also interact most likely with phosphate oxygens of POPC, 
but in contrast to calcium, the interactions with only carbonyls are also significant.
The physiological relevance of the sodium binding details is, however, uncertain
due to its very weak binding affinity.
